---
title: "R Programming for Time Series Analysis of Optimized Natural Gas Portfolios"
author: "Lewis Boslego"
date: "March 14, 2023"
output:
  pdf_document: default
---



```{r}
library(tidyverse)
library(magrittr) # double pipe
library(forecast) # time series functions
library(MASS) # boxcox
library(tseries) #adf.test()
library(gridExtra) # graphing tool
library(knitr)
```

Import data:
```{r}
ng <- read_csv("174_data.csv", show_col_types = "f")
ng %>% head()

# The columns with ticker names are calender month arithmetic averages of 
# the stocks daily adjusted close prices.
# the columns with ticker names followed by _return are the logarithmic averages
# of the monthly average price. Formula: Rt = Ln(Pt/Pt-1)
# All of this data processing was done using Microsoft Excel because it is far
# better suited for these steps. 

# NOTE: Date is inaccurate, all data is end of month average. 
# e.g. observations for June 1, 2000, actually correspond to June 30, 2000

ng_train <- ng[1:282,] # data to train models, removed last 12 months.
ng_test <- ng[283:294,] # only most recent 12 months of data to test forecasts.
```



---
Generating Equal Weights (EW):
```{r}
portfolio <- ng_train 
portfolio %<>% mutate(equal_port_price = .2*EQT + .2*DVN + 
                                         .2*LNG + .2*EPD + 
                                         .2*SRE, 
                      equal_port_return = .2*EQT_return + .2*DVN_return + 
                                          .2*LNG_return + .2*EPD_return + 
                                          .2*SRE_return) 
```

Generating Global Minimum Variance (GMV):
```{r}
cov_mtx <- portfolio %>% 
           dplyr::select(EQT_return, DVN_return, 
                         LNG_return, EPD_return, 
                         SRE_return)  %>%  cov()

gmv_cov_mtx <- rbind(2*cov_mtx, c(1, 1, 1, 1, 1))
gmv_cov_mtx %<>% cbind(c(1, 1, 1, 1, 1, 0))
b <- c(0, 0, 0, 0, 0, 1)

w <- solve(gmv_cov_mtx)%*%b
w_vec <- w[1:5, 1]
w_vec # Global Minimum Variance weights

portfolio %<>% mutate(mv_port_price = 
                       0.1703922*EQT + -0.1111337*DVN + 
                      -0.0194195*LNG +  0.4235468*EPD +
                       0.5366141*SRE,
                       mv_port_return = 
                        0.1703922*EQT_return + -0.1111337*DVN_return + 
                       -0.0194195*LNG_return +  0.4235468*EPD_return + 
                        0.5366141*SRE_return)
```


Maximum Sharpe Ratio (SRP):
```{r}
mean_return <- c(mean(ng_train$EQT_return), mean(ng_train$DVN_return),
                 mean(ng_train$LNG_return), mean(ng_train$EPD_return),
                 mean(ng_train$SRE_return))

sd_return <- c(sd(ng_train$EQT_return), sd(ng_train$DVN_return),
               sd(ng_train$LNG_return), sd(ng_train$EPD_return),
               sd(ng_train$SRE_return))

sigma_inv_mat <- cov_mtx %>% solve()
one_vec <- rep(1,5)

top_mat = sigma_inv_mat%*%mean_return[1:5]
bot_val = as.numeric(t(one_vec)%*%top_mat)
m_vec = top_mat[,1]/bot_val
m_vec # maximum Sharpe portfolio weights

portfolio %<>% mutate(sharpe_port_price = 
                       0.134725966*EQT + -0.260390861*DVN +
                      -0.001451409*LNG +  0.701833301*EPD + 
                       0.425283002*SRE,
                     sharpe_port_return = 
                        0.134725966*EQT_return + -0.260390861*DVN_return +
                       -0.001451409*LNG_return + 0.701833301*EPD_return + 
                        0.425283002*SRE_return) 

Sharpe = mean(portfolio$sharpe_port_return) / 
         sd(portfolio$sharpe_port_return) # = 0.245
```

Preparing data for mean-variance graph.
```{r}
mean_return %<>% append(
                      c(mean(portfolio$equal_port_return), 
                        mean(portfolio$mv_port_return), 
                        mean(portfolio$sharpe_port_return)))

sd_return %<>% append(
                    c(sd(portfolio$equal_port_return), 
                      sd(portfolio$mv_port_return), 
                      sd(portfolio$sharpe_port_return)))
               
names <- c('EQT','DVN','LNG', 'EPD', 'SRE', 'EW', 'GMV', 'SRP')
# EW is equal weights, GMV is global minimum variance, SRP is maximum Sharpe 
# ratio.

mean_sd <- cbind(names, mean_return, sd_return) %>% 
           as_tibble() %>% 
           mutate(mean_return = as.numeric(mean_return), 
                    sd_return = as.numeric(sd_return))
```


Calculating the portfolios for the testing data set
```{r, include=F}
portfolio_test <- ng_test %>% mutate(
                      equal_port_return = 
                          .2*EQT_return + .2*DVN_return + 
                          .2*LNG_return + .2*EPD_return + 
                          .2*SRE_return,
                       mv_port_return = 
                 0.1703922*EQT_return + -0.1111337*DVN_return + 
                -0.0194195*LNG_return +  0.4235468*EPD_return + 
                 0.5366141*SRE_return,
                 sharpe_port_return = 
             0.134725966*EQT_return + -0.260390861*DVN_return +
            -0.001451409*LNG_return + 0.701833301*EPD_return + 
             0.425283002*SRE_return)

portfolio_full <- ng %>% mutate(
                      equal_port_return = 
                          .2*EQT_return + .2*DVN_return + 
                          .2*LNG_return + .2*EPD_return + 
                          .2*SRE_return,
                       mv_port_return = 
                 0.1703922*EQT_return + -0.1111337*DVN_return + 
                -0.0194195*LNG_return +  0.4235468*EPD_return + 
                 0.5366141*SRE_return,
                 sharpe_port_return = 
             0.134725966*EQT_return + -0.260390861*DVN_return +
            -0.001451409*LNG_return + 0.701833301*EPD_return + 
             0.425283002*SRE_return)
```

Plot of all 3 portfolios and all stocks on mean-variance plane
```{r}
ggplot(data = mean_sd, aes(x=sd_return, y=mean_return)) + 
 geom_point() +
  geom_text(aes(label = names), vjust = 1.45, color = "blue", cex = 3.5) +
  labs(x = 'Volatility (sd)', y = 'Average Return', 
       title = "Mean-Variance Representation of Market Basket",
       subtitle = "Individual stocks and weighted portfolios") + 
  theme_bw() 
```

Graph of prices
```{r}
par(mfrow = c(2,2))
plot(y =ng_train$EQT, x = ng_train$date, type = "l",
     main = "EQT", ylab = "Price",  xlab = "Date")
plot(y = ng_train$DVN, x = ng_train$date, type = "l",
     main = "DVN", ylab = "Price",  xlab = "Date")
plot(y =ng_train$LNG,x = ng_train$date, type = "l",
     main = "LNG", ylab = "Price",  xlab = "Date")
plot(y = ng_train$EPD, x = ng_train$date, type = "l",
     main = "EPD", ylab = "Price",  xlab = "Date")
plot(y = ng_train$SRE, x = ng_train$date, type = "l", 
     main = "SRE", ylab = "Price",  xlab = "Date")
```

Graph of returns
```{r}
par(mfrow = c(2,2))
plot(y =ng_train$EQT_return, x = ng_train$date, type = "l",
     main = "EQT", ylab = "Return",  xlab = "Date")
plot(y = ng_train$DVN_return, x = ng_train$date, type = "l",
     main = "DVN", ylab = "Return",  xlab = "Date")
plot(y =ng_train$LNG_return,x = ng_train$date, type = "l",
     main = "LNG", ylab = "Return",  xlab = "Date")
plot(y = ng_train$EPD_return, x = ng_train$date, type = "l",
     main = "EPD", ylab = "Return",  xlab = "Date")
plot(y = ng_train$SRE_return, x = ng_train$date, type = "l", 
     main = "SRE", ylab = "Return",  xlab = "Date")
```


ARIMA:
---
Equal weight
---
Price and Return plot:
```{r}
a1 <- ggplot(portfolio, aes(x = date, y = equal_port_price)) + geom_line() + 
  labs(y = 'Price',x = "Date") + theme_minimal()
b1 <- ggplot(portfolio, aes(x = date, y = equal_port_return)) + geom_line() + 
  labs(y = 'Return', x = "Date") + theme_minimal()
gridExtra::grid.arrange(a1, b1, top = "Equal Weight Portfolio")
```

stationary test:
```{r}
portfolio$equal_port_return %>% ts() %>% 
  adf.test(alternative = "stationary") 
# H0: non-stationary. p < .01 => reject H0.
```


ACF and PACF
```{r}
par(mfrow = c(2,1))
acf(portfolio$equal_port_return, main = "Equal Weight Portfolio Autocorrelations")
pacf(portfolio$equal_port_return, main = "") 
```


MA(1)
```{r}
equal_ma1 <- forecast::Arima(portfolio$equal_port_return, 
                             order = c(0, 0, 1), method = "ML")
summary(equal_ma1)
```

AR(1)
```{r, echo=F}
equal_ar1 <- forecast::Arima(portfolio$equal_port_return, 
                             order = c(1, 0, 0), method = "ML")
summary(equal_ar1)
```

ARMA(1,1)
```{r}
equal_arma11 <- forecast::Arima(portfolio$equal_port_return, 
                             order = c(1, 0, 1), method = "ML")
summary(equal_arma11)
```

```{r}
checkresiduals(equal_ar1, plot = F)

checkresiduals(equal_ma1, plot = F)

checkresiduals(equal_arma11, plot = F)
```

Using the models to forecast:
```{r}
equal_ma1_forecast <- forecast::forecast(equal_ma1, h=12) 
equal_ar1_forecast <- forecast::forecast(equal_ar1, h=12) 
equal_arma11_forecast <- forecast::forecast(equal_arma11, h=12) 
```

```{r}
par(mfrow = c(2,1))
plot(equal_ma1_forecast, 
     xlim = c(length(portfolio$equal_port_return)-36, 
              length(portfolio$equal_port_return)+12))
plot(equal_ar1_forecast, 
     xlim = c(length(portfolio$equal_port_return)-36, 
              length(portfolio$equal_port_return)+12))
plot(equal_arma11_forecast, 
     xlim = c(length(portfolio$equal_port_return)-36, 
              length(portfolio$equal_port_return)+12)) 
plot(ts(portfolio_full$equal_port_return), xlim = 
       c(length(portfolio_full$equal_port_return)-48, 
              length(portfolio_full$equal_port_return)),
     main = "True Return Time Series", xlab="",ylab="")
```


Error statistics for forecast
```{r}
# accuracy of model predictions against the testing data.
equal_ma1_accuracy <- forecast::accuracy(equal_ma1_forecast, 
                                         ts(portfolio_full$equal_port_return))
equal_ar1_accuracy <- forecast::accuracy(equal_ar1_forecast, 
                                         ts(portfolio_full$equal_port_return)) 
equal_arma11_accuracy <- forecast::accuracy(equal_arma11_forecast,
                                          ts(portfolio_full$equal_port_return)) 


equal_test_rmse <- c(round(equal_ma1_accuracy[2,2], digits = 5),
               round(equal_ar1_accuracy[2,2], digits=5),
               round(equal_arma11_accuracy[2,2], digits=5))
equal_test_mae <- c(round(equal_ma1_accuracy[2,3], digits=5),
               round(equal_ar1_accuracy[2,3], digits =5),
               round(equal_arma11_accuracy[2,3], digits=5))
equal_test_names <- c("MA(1)","AR(1)","ARMA(1,1)")

equal_tests <- cbind(equal_test_names,equal_test_rmse,equal_test_mae) %>% as_data_frame()

kable(
  equal_tests %>% rename("Model" = "equal_test_names", 
                         "RMSE" = "equal_test_rmse",
                         "MAE" = "equal_test_mae"),
  caption = "Model Testing Accuracy")
```

gmv:
---
gmv price & return plot: 
```{r}
a2 <- ggplot(portfolio, aes(x = date, y = mv_port_price)) + geom_line() + 
  labs(y = 'Price',x = "Date") + theme_minimal()
b2 <- ggplot(portfolio, aes(x = date, y = mv_port_return)) + geom_line() + 
  labs(y = 'Return', x = "Date") + theme_minimal()
gridExtra::grid.arrange(a2, b2, top = "Global Minimum Variance")
```

stationary test
```{r}
adf.test(portfolio$mv_port_return) # stationary
```

ACF & PACF gmv:
```{r}
acf(portfolio$mv_port_return, main = "")  
pacf(portfolio$mv_port_return, main = "") 
```

Fitting the models
```{r}
gmv_ma3 <- forecast::Arima(portfolio$mv_port_return, order = c(0,0,3), method = "ML")
summary(gmv_ma3)
gmv_ma1 <- forecast::Arima(portfolio$mv_port_return, order = c(0,0,1), method = "ML")
summary(gmv_ma1)
gmv_ma0 <- forecast::Arima(portfolio$mv_port_return, order =c(0,0,0), method = "ML")
summary(gmv_ma0)
```

Checking residuals:
```{r}
checkresiduals(gmv_ma3, plot = F)
checkresiduals(gmv_ma1, plot = F)
checkresiduals(gmv_ma0, plot = F)
```

```{r, echo = F}
gmv_ma3_forecast <- forecast::forecast(gmv_ma3, h=12) 
gmv_ma1_forecast <- forecast::forecast(gmv_ma1, h=12) 
gmv_ma0_forecast <- forecast::forecast(gmv_ma0, h=12) 
```

Generating forecasts
```{r}
par(mfrow = c(2,1))
plot(gmv_ma3_forecast, 
     xlim = c(length(portfolio$mv_port_return)-36, 
              length(portfolio$mv_port_return)+12))
plot(gmv_ma1_forecast, 
     xlim = c(length(portfolio$mv_port_return)-36, 
              length(portfolio$mv_port_return)+12))
plot(gmv_ma0_forecast, 
     xlim = c(length(portfolio$mv_port_return)-36, 
              length(portfolio$mv_port_return)+12)) 
plot(ts(portfolio_full$mv_port_return), xlim = 
       c(length(portfolio_full$mv_port_return)-48, 
              length(portfolio_full$mv_port_return)),
     main = "True Return Time Series", xlab="",ylab="")
```

# forecasting error
```{r}
# accuracy of model predictions against the testing data.
gmv_ma3_accuracy <- forecast::accuracy(gmv_ma3_forecast, 
                                         ts(portfolio_full$mv_port_return))
gmv_ma1_accuracy <- forecast::accuracy(gmv_ma1_forecast, 
                                         ts(portfolio_full$mv_port_return)) 
gmv_ma0_accuracy <- forecast::accuracy(gmv_ma0_forecast,
                                          ts(portfolio_full$mv_port_return)) 


gmv_test_rmse <- c(round(gmv_ma3_accuracy[2,2], digits = 5),
               round(gmv_ma1_accuracy[2,2], digits=5),
               round(gmv_ma0_accuracy[2,2], digits=5))
gmv_test_mae <- c(round(gmv_ma3_accuracy[2,3], digits=5),
               round(gmv_ma1_accuracy[2,3], digits =5),
               round(gmv_ma0_accuracy[2,3], digits=5))
gmv_test_names <- c("MA(3)","MA(1)","MA(0)")

gmv_tests <- cbind(gmv_test_names, gmv_test_rmse, gmv_test_mae) %>% 
             as_data_frame()

kable(
  gmv_tests %>% rename("Model" = "gmv_test_names", 
                         "RMSE" = "gmv_test_rmse",
                         "MAE" = "gmv_test_mae"),
  caption = "Model Testing Accuracy")
```




Sharpe portfolio:
---
```{r}
a3 <- ggplot(portfolio, aes(x = date, y = sharpe_port_price)) + 
  geom_line() + labs(y = 'Price',x = "Date") + theme_minimal()
b3 <- ggplot(portfolio, aes(x = date, y = sharpe_port_return)) + 
  geom_line() + labs(y = 'Return', x = "Date") + theme_minimal()
gridExtra::grid.arrange(a3, b3, top = "Maximum Sharpe Ratio Portfolio")
```

stationary test:
```{r}
adf.test(portfolio$sharpe_port_return, alternative = "stationary")
```

ACF & PACF sharpe:
```{r}
acf(portfolio$sharpe_port_return, main = "")  
pacf(portfolio$sharpe_port_return, main = "") 
```

Fitting models:
```{r}
sharpe_ma3 <- forecast::Arima(portfolio$sharpe_port_return,
                              order = c(0,0,3), 
                              method = "ML")
#summary(sharpe_ma3)
sharpe_ma1 <- forecast::Arima(portfolio$sharpe_port_return,
                              order = c(0,0,1), 
                              method = "ML")
# summary(sharpe_ma1)
sharpe_ar3 <- forecast::Arima(portfolio$sharpe_port_return,
                              order = c(3,0,0), 
                              method = "ML")
# summary(sharpe_ar3)
```

checking model residuals:
```{r}
checkresiduals(sharpe_ma3, plot = F) # good 
checkresiduals(sharpe_ma1, plot = F) # bad 
checkresiduals(sharpe_ar3, plot = F) # good
```

forecasting models:
```{r}
sharpe_ma3_forecast <- forecast::forecast(sharpe_ma3, h=12) 
sharpe_ar3_forecast <- forecast::forecast(sharpe_ar3, h=12) 

par(mfrow = c(2,1))
plot(sharpe_ma3_forecast, 
     xlim = c(length(portfolio$sharpe_port_return)-36, 
              length(portfolio$sharpe_port_return)+12))
plot(sharpe_ar3_forecast, 
     xlim = c(length(portfolio$sharpe_port_return)-36, 
              length(portfolio$sharpe_port_return)+12))
plot(ts(portfolio_full$sharpe_port_return), xlim = 
       c(length(portfolio_full$sharpe_port_return)-48, 
              length(portfolio_full$sharpe_port_return)),
     main = "True Return Time Series", xlab="",ylab="")
```

model accuracy:
```{r}
# accuracy of model predictions against the testing data.
sharpe_ma3_accuracy <- forecast::accuracy(sharpe_ma3_forecast, 
                                         ts(portfolio_full$sharpe_port_return))
sharpe_ar3_accuracy <- forecast::accuracy(gmv_ma1_forecast, 
                                         ts(portfolio_full$sharpe_port_return)) 



sharpe_test_rmse <- c(round(sharpe_ma3_accuracy[2,2], digits = 5),
               round(sharpe_ar3_accuracy[2,2], digits=5))
sharpe_test_mae <- c(round(sharpe_ma3_accuracy[2,3], digits=5),
               round(sharpe_ar3_accuracy[2,3], digits =5))
sharpe_test_names <- c("MA(3)","AR(3)")

sharpe_tests <- cbind(sharpe_test_names, sharpe_test_rmse, sharpe_test_mae) %>% 
             as_data_frame()

kable(
  sharpe_tests %>% rename("Model" = "sharpe_test_names", 
                         "RMSE" = "sharpe_test_rmse",
                         "MAE" = "sharpe_test_mae"),
  caption = "Model Testing Accuracy")
```







